Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2023-08-30 manuscript_thev_transcriptome.Rmd
Ballgown trxpt expression analysis and figs

## RESULTS   
**Overview of sequencing data and analysis pipeline outputs**     
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `HISAT2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (__Table 1__, __Figure 2a__). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our qPCR assay quantifying the total number of viral genome copies over time (__Figure 2b__). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the final consolidated transcriptome of THEV, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (__Figure 3__). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (http://www.ncbi.nlm.nih.gov/geo) under **accession no. XXXXXX**.

**Changes in THEV splice junction profile over time**   
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU). Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs. To evaluate the activity of each promoter over time, we estimate relative abundances of all splice junctions for each time point, counting as significantly expressed only those with coverage of at least 1% of the total splice junction reads counted at the given time point. Very few unique splice junctions and reads were counted at 4 h.p.i; hence, this time point was excluded in this analysis. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were predominantly early region (E1, E2, and E3) junctions, albeit a significant proportion are MLP-derived (__Table 2a__). Junctions from the E2 region were the most abundant at this time point, constituting `r round(most_sig_12, 1)`% of all junctions. The two most abundant junctions at 12 h.p.i were maintained at 24 h.p.i also as the most significantly expressed. However, the MLP-derived junctions were unsurprisingly the most preponderant overall at this time point, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (__Table 2b__). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions was not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (__Table 2c__. Also see __Supplementary Tables 1a-c__; __Figure 4a__). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was substantially similar to the pattern seen with all the junctions included (__Figure 4b__).

Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. This revealed that, with the exception of one junction (which was a T-G donor-acceptor pair), all splice donor-acceptor pairs in the final transcriptome were the canonical T-A pair.

@@

![](results/r/figures/junc_abundances.png)
**Figure 4. a) Relative abundances of splice junctions over time as a proxy of estimating activity levels of THEV promoters**. ## Description here ##########. **b) Relative abundances of junctions in final transcriptome**. ##Description here #######.

>>>>

## RESULTS   
**Overview of sequencing data and analysis pipeline outputs**     
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (__Table 1__, __Figure 2a__). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our qPCR assay quantifying the total number of viral genome copies over time (__Figure 2b__). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the final consolidated transcriptome of THEV, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (__Figure 3__). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (http://www.ncbi.nlm.nih.gov/geo) under **accession no. XXXXXX**.

**Changes in THEV splice junction profile over time**   
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU). Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs. To evaluate the activity of each promoter over time, Firstly, `Ballgown` - a program for statistical analysis of assembled transcriptomes, was used to estimate and normalize expression levels of all transcripts in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units over time. This revealed DBP - from the E2 region - to be the most significantly expressed at 12 h.p.i, whereas MLP region transcripts were most abundantly expressed at 24 and 72 h.p.i as expected (__Figure 4a__). Also, our analysis of the FPKM values of transcripts by region showed a similar pattern: the E2 region as the most significantly overexpressed at 12 h.p.i, after which the MLP region assumes predominance (__Figure 4b__).
Secondly, we estimated relative abundances of all splice junctions for each time point, counting as significantly expressed only those with coverage of at least 1% of the total splice junction reads counted at the given time point. Very few unique splice junctions and reads were counted at 4 h.p.i; hence, this time point was excluded in this analysis. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were predominantly early region (E1, E2, and E3) junctions, albeit a significant proportion are MLP-derived (__Table 2a__). Junctions from the E2 region were the most abundant at this time point, constituting `r round(most_sig_12, 1)`% of all junctions. The two most abundant junctions at 12 h.p.i were maintained at 24 h.p.i also as the most significantly expressed. However, the MLP-derived junctions were unsurprisingly the most preponderant overall at this time point, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (__Table 2b__). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions was not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (__Table 2c__. Also see __Supplementary Tables 1a-c__; __Figure 5a__). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was substantially similar to the pattern seen with all the junctions included (__Figure 5b__).
Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. This revealed that, with the exception of one junction (which was a T-G donor-acceptor pair), all splice donor-acceptor pairs in the final transcriptome were the canonical T-A pair.

@@

![](results/r/figures/trxpt_fpkm_percent_abund.png)

![](results/r/figures/region_fpkm_percent_abund.png)
**Figure 4 Overview of change in expression of THEV transcripts over time.** 
**a)** 
**b)**

![](results/r/figures/junc_abundances.png)
**Figure 5. a) Relative abundances of splice junctions over time as a proxy of estimating activity levels of THEV promoters**. ## Description here ##########. **b) Relative abundances of junctions in final transcriptome**. ##Description here #######.
